### 2017-04-06 paper/sexSelectionSpacing-ver1.tex
Further edits of theory part; moved tables out of doc; moved comparison back to appendix

The males-to-females ratio at birth increased dramatically over the last three 
decades in India as access to prenatal sex determination expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth \citep{Yi1993,park95}.
}
Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask about use of 
sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have relied on a simple method for establishing 
factors that affect sex selection: use the sex of children born as the 
dependent variable, and estimate the effects of variables on the probability of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

@@

@@

It also highlights the need to incorporate fertility progression into
estimates, something the simple model is incapable of.

@@

% Highlights and tie into empirical model:

An important question here is whether parents ever
change their use of prenatal sex determination between
two consecutive births.
I capture this by the percent of spells where prenatal
scans were initially used for one or more period, but the birth
is the result of a pregnancy without a prenatal scan.
Whether parents will consider the duration from last birth too long 
depends mainly on the discount rate.
On average, the more impatient parents are---the less they
like long spells between births---the more likely they are to
change their decision to use sex selection.%
\footnote{
Using only those situations where there are any abortions,
regressing the discount rate against the percent spells where
prenatal scan is initially used but not for the birth show
a strong positive effect---higher number, more patient---holding
all the other parameters constant.
}
Lowering the cost of prenatal sex determination generally leads to 
a smaller percentage of parents who change their use of sex selection.
But, the total number of spells with a change may still
go up because lower cost of prenatal sex determination also
increases the use of sex selection.

% What are the model highlights 
There are three main conclusions from the model.%
\footnote{
A caveat is that these results reflect the particular utility 
function chosen.
If parents have a different utility function, like a preference
for one son but otherwise no preference across boys and girls
the results may look different.
}
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

@@

The theoretical model has three implications for fertility decisions with access to 
sex-selective abortions: 
a higher probability that next child is a son, 
longer average time to next birth, 
and that the use of prenatal sex determination may change between one observed 
birth and the next.
All three must be incorporated into the estimation strategy
to provide unbiased and precise estimates of the use of sex-selective abortions.%

@@

@@

The data come from the three rounds of the National Family Health Survey 
(NFHS-1, NFHS-2 and NFHS-3),
collected in 1992--93, 1998--99, and 2005--2006.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has a small number of observation 
collected in 2000.
}
The International Institute for Population Sciences in Mumbai collected the data,
which have nationwide coverage.
% All are based on the Demographic and Health Survey Model B, with
% NFHS-1 using the DHS II questionnaire while NFHS-2 and NFHS-3 using, 
% respectively, the DHS III questionnaire and the Measure DHS questionnaire.
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households,
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households
and NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households.

@@

By dividing spells into these periods I can examine how fertility
decisions and the use of sex selection have changed across the three different regimes.
Note that the periods are based on the spells' beginning year, and some spells 
will therefore cover two periods.
A couple may, for example, be married in 1984, but not have their first child until 1986.
That couple's first spell will be in the 1972--1984 period, even though most of the 
spell actually falls in the 1985--1994 period.
Some children born from spells that began in the 1972--1984 period may therefore have been
conceive when prenatal sex determination techniques were available, which could result
in evidence of sex-selective abortions even for this period.
Similarly, a spell that began in the 1985--1994 period may have been partly or mostly
under the PDNT act.
The overall effect is to bias downwards any differences between the periods.

@@

\subsection{Comparing the Hazard and Simple Models}

As detailed in the main paper, one of the main advantages of the hazard model 
approach is that it can be used to predict what completed fertility, number of abortions, 
and sex ratios will be once women are done with childbearing.
This cannot be done with the simple model because it does not predict fertility
progression, does not take into account censoring, and cannot capture if parents
change their use of sex selection within a spell.
We can, however, still compare differences in predicted sex ratios for 
\emph{individual} parities.%
\footnote{
The simple model uses only observed births and the predicted sex ratio is 
therefore simply what we observe for the relevant uncensored sample.
}

For each parity, the simple model is
\begin{equation}
Y_i = \gamma + \alpha' \mathbf{Z_i} + \beta' \mathbf{X}_{i} + \epsilon_i,
\end{equation}
where $Y$ is a dummy variable that takes the value 1 if the child born is a boy and zero 
if it is a girl, and the vectors of explanatory variables, $\mathbf{Z}_{i}$ and 
$\mathbf{X}_{i}$, are the same as for the proposed method, except that 
$\mathbf{Z}_{i}$ obviously does not include the baseline hazard and that there are 
no time dependence.%
\footnote{
To ease presentation, the indicator for parity number is not shown.
}
I estimate this using a logit model, although a linear probability model leads to
similar results.
This model is estimated on all observed births for a given parity in the relevant 
period to ensure that the results are as close to the standard method of 
examining sex selection \citep[see, for example, ][]{retherford03b,jha06,abrevaya09}.

\begin{table}[htbp]
\begin{center}
\begin{footnotesize}
\begin{threeparttable}
\caption{Comparison of Hazard and Simple Models for  \protect\linebreak
Women with 8 or More Years of Education}
\label{tab:compare2}
\begin{tabular} {@{} l D{.}{.}{5.2} D{.}{.}{5.2} D{.}{.}{5.2} D{.}{.}{5.2}   @{}} \toprule
                                  &  \multicolumn{2}{c}{Urban}        & \multicolumn{2}{c}{Rural}         \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
                                  & \mco{Simple}     &  \mco{Hazard}   &  \mco{Simple}    & \mco{Hazard}      \\   \midrule
                                  & \multicolumn{4}{c}{1985--1994 - Second Birth} \\ \cmidrule(l){2-5}
                                  & \multicolumn{4}{c}{1 girl} \\
Number of women\tnote{a}          &             & 4,869   &         & 3,105   \\
Births\tnote{b}                   &     3,848   & 3,926   & 2,417   & 2,795   \\
Percentage boys                   &      54.0   &  54.7   &  52.6   &  53.5   \\
Sex ratio (boys per 100 girls)    &     117.4   & 120.7   & 110.9   & 115.1   \\
\addlinespace 
                                  & \multicolumn{4}{c}{1985--1994 - Third Birth}  \\ \cmidrule(l){2-5}
%                                   & \multicolumn{4}{c}{1 boy and 1 girl}        \\
% Number of women\tnote{a}          &             & 3,460   &         & 1,987   \\
% Births\tnote{c}                   &     1,431   & 1,558   & 1,107   & 1,336   \\
% Percentage boys                   &      53.0   &  52.5   &  53.5   &  54.6   \\
% Sex ratio (boys per 100 girls)    &     112.9   & 110.4   & 115.0   & 120.2   \\
% \addlinespace
                                  & \multicolumn{4}{c}{2 girls}        \\
Number of women\tnote{a}          &             & 1,654   &         &   982   \\
Births\tnote{c}                   &     1,034   & 1,092   &   682   &   853   \\
Percentage boys                   &      60.8   &  63.7   &  54.7   &  57.4   \\
Sex ratio (boys per 100 girls)    &     155.3   & 175.3   & 120.7   & 134.6   \\
\addlinespace 
                                  & \multicolumn{4}{c}{1995--2006 - Second Birth} \\ \cmidrule(l){2-5}
                                  & \multicolumn{4}{c}{1 girl} \\                                                  
Number of women\tnote{a}          &             & 3,774   &         & 2,823   \\
Births\tnote{b}                   &     3,031   & 2,704   & 2,489   & 2,407   \\
Percentage boys                   &      58.0   &  58.6   &  56.0   &  56.9   \\
Sex ratio (boys per 100 girls)    &     138.1   & 141.3   & 127.1   & 132.2   \\
\addlinespace 
                                  & \multicolumn{4}{c}{1995--2006 - Third Birth}  \\ \cmidrule(l){2-5}
%                                   & \multicolumn{4}{c}{1 boy and 1 girl}        \\
% Number of women\tnote{a}          &             & 2,357   &         & 1,819   \\
% Births\tnote{c}                   &       830   &   775   &   853   &   998   \\
% Percentage boys                   &      53.9   &  52.9   &  55.5   &  56.5   \\
% Sex ratio (boys per 100 girls)    &     116.7   & 112.1   & 124.5   & 129.8   \\
% \addlinespace
                                  & \multicolumn{4}{c}{2 girls}        \\
Number of women\tnote{a}          &             & 1,000   &         &   863   \\
Births\tnote{c}                   &       695   &   600   &   690   &   661   \\
Percentage boys                   &      65.5   &  62.4   &  62.0   &  61.7   \\
Sex ratio (boys per 100 girls)    &     189.6   & 166.0   & 163.4   & 161.0   \\
\bottomrule
\end{tabular}                        
\begin{tablenotes} \tiny
\item \hspace*{-0.5em} \textbf{Note.} 
The simple models are estimated using logit on all births for a given parity that
occurred during the relevant time period.
The hazard models are estimated using all spells that began in the relevant time period.
For the second birth the hazard model covers the period from beginning of spell to 6 
years (24 quarters) after the birth of the first child.
For the third birth it covers the period from beginning of spell to 7.25 years 
(29 quarters) after the birth of the second child for the period 1985--1994 and from 
beginning of spell to 5.75 years (23 quarters) after the birth of the second child for 
the period 1995--2006.
\item[a] Number of women in period who began the spell with given sex composition of prior 
child/children in the relevant period.
\item[b] For Logit model the number of women who are observed to have a second birth
in the period. 
For hazard model the predicted number of second births that will occur between beginning 
of the spell and 6 years (24 quarters) after the birth of the first child.
% \item[c] Number of women in period who had one boy and one girl as their first two children.
\item[c] For Logit model the number of women who are observed to have a third birth
in the period. 
For hazard model the predicted number of third births that will occur between beginning of 
the spell and 7.25 years (29 quarters) after the birth of the second child for the
1985--1994 period and between beginning of the spell and 5.75 years (23 quarters) after 
the birth of the second child for the 1995--2006 period.
\end{tablenotes}
\end{threeparttable}
\end{footnotesize}
\end{center}
\end{table}

I focus on those cases where there is evidence of sex-selective abortions: the spells
from parity one to two and two to three for women with eight or more years of education.%
\footnote{
The results for the spell from parity three to four are available on request.
They are not presented because the number of educated women who end up with three births 
and the probability of having a fourth birth are small, which makes
comparing the simple and hazard models with any precision difficult.
}
Table \ref{tab:compare2} shows the number of women who enter the second and third spells
for the hazard model,
the number of births, observed in the case of the simple model and predicted in the case 
of the hazard model, predicted percent boys, and the associated predicted sex ratio.
It is possible for the number of births used for the simple model to be larger than the
hazard models' predicted number of births because the simple model is based
on all births that occurred in the period---independently of when the prior birth
occurred---whereas the hazard model uses only spells that began in the period.

For the 1985--1994 period---which is when sex selection become available---the 
hazard model consistently predicts a higher sex ratio than the logit model.
Furthermore, some of these differences are substantial.
For urban women with 2 girls as their first two children, the hazard model
predicts a sex ratio of 175 boys per 100 girls, whereas the simple method
``only'' predicts 155 boys per 100 girls.
The difference is smaller in rural areas but still 14 boys per 100 girls
higher for the hazard model than for the simple model.
A similar, if less pronounced pattern, show up for the second birth,
where the hazard model predict 3 and 5 boys more per 100 girls for
urban and rural women.

For the latest period, 1995--2006, the hazard model still produces higher 
predicted sex ratios for second birth and the differences are of a 
similar magnitude to the 1985--1994 period.
The predicted sex ratios for the third birth are, however, 
higher for the simple model than for the hazard model.
This difference is substantially for urban women where the difference 
is 23 boys extra per 100 girls.

A possible explanation for these differences is that the simple model 
fails to capture the move towards lower fertility and the subsequent 
increase in the use of sex-selective abortions at earlier parities that 
occurred over the periods.
The results also suggest that the take-up of sex selection occurred 
rapidly as the new methods became available and that this take-up
is not fully captured by the simple model.
More broadly, a possible interpretation of the differences is that
the simple model will tend to underestimate when sex-selective abortions
are initially spreading in use and overestimate when the spread of sex 
selection is slowing.
This is most clearly seen for the third births for urban women where
the hazard model show a \emph{decline} is use of sex-selective abortions,
consistent with the lower fertility as discussed in the main paper,
while the simple model show a substantial increase.

@@

In the discussion of declining fertility I rely on sex selection not being used by 
families that already has one son.
The extent to which this holds is of interest for two reasons.
First and foremost, it can help us understand what form son preference takes in India.
This is important in itself, but especially as we try to predict what will happen to 
fertility and sex selection use in the future.
Second,
if families use sex selection to reach two or more sons, progression rates for families 
with one son may overestimate the decline in fertility.

@@

\begin{table}[htbp]
\begin{center}
\begin{scriptsize}
\begin{threeparttable}
\caption{Predicted Fertility Behavior, Sex Ratio, and \protect\linebreak
Abortions for Women with 8 or More Years of Education}
\label{tab:predicted}
\begin{tabular} {@{} l D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1}  D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1} D{.}{.}{2.1}  @{}} \toprule
                 & \multicolumn{4}{c}{1972-1984}                                                               & \multicolumn{4}{c}{1985-1994}                                                               & \multicolumn{4}{c}{1995-2006} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(lr){10-13}
                 & \mco{Births (\%)}          & \mct{Abortions per 100}                      & \mco{Boys}      & \mco{Births (\%)}          & \mct{Abortions per 100}                      & \mco{Boys}      & \mco{Births (\%)}          & \mct{Abortions per 100}                      & \mco{Boys}      \\
                 & \mco{/ fertility\tnote{a}} & \mco{births\tnote{b}} & \mco{women\tnote{c}} & \mco{(\%)}      & \mco{/ fertility\tnote{a}} & \mco{births\tnote{b}} & \mco{women\tnote{c}} & \mco{(\%)}      & \mco{/ fertility\tnote{a}} & \mco{births\tnote{b}} & \mco{women\tnote{c}} & \mco{(\%)}      \\
\midrule
\bf Urban        & \multicolumn{12}{c}{} \\                                                                    
Spell 2          &       85.9   &  2.0    &  1.8    &  52.3   &       78.2   &  4.7    &  3.7  &  53.7   &       68.0   &  8.0    &  5.4   &  55.3      \\
Spell 3          &       64.4   &  5.1    &  3.3    &  53.9   &       49.9   &  9.8    &  4.9  &  56.2   &       38.6   &  7.8    &  3.0   &  55.3      \\
Spell 4          &       57.0   &  2.8    &  1.6    &  52.6   &       44.8   &  2.6    &  1.1  &  52.6   &       33.7   &  21.8   &  7.3   &  62.3      \\
Overall\tnote{d} &       2.7    &  2.0    &  5.5    &  52.3   &       2.3    &  3.4    &  8.0  &  53.1   &       2.0    &  4.6    &  9.4   &  53.7      \\
\addlinespace 
\bf Rural        & \multicolumn{12}{c}{} \\                                                                                  
Spell 2          &       92.5   &  -1.4   &  -1.3   &  50.5   &       87.6   &  3.1    &  2.7  &  52.8   &       83.1   &  4.4    &  3.6   &  53.5      \\
Spell 3          &       80.9   &  2.6    &  2.1    &  52.5   &       70.6   &  6.5    &  4.6  &  54.5   &       59.6   &  10.6   &  6.3   &  56.7      \\
Spell 4          &       75.7   &  -3.1   &  -2.4   &  49.6   &       62.0   &  0.2    &  0.1  &  51.3   &       51.9   &  6.1    &  3.2   &  54.4      \\
Overall\tnote{d} &       3.2    &  -0.3   &  -1.1   &  51.2   &       2.9    &  2.4    &  6.8  &  52.8   &       2.6    &  4.1    &  10.5  &  53.4      \\
\bottomrule
\end{tabular}                        
\begin{tablenotes} \tiny
\item \hspace*{-0.5em} \textbf{Note.} Predictions are based on estimates of equations 
(\ref{eq:probability_boy}) and (\ref{eq:probability_girl}).
The samples consist of all women from the data with 8 or more years of education who
married during the indicated period using the estimation results for that periods 
and the women's characteristics at the start of their marriage. 
All numbers presented are for women predicted to have one or more births.
Birth/progression rates are based on the ratio of women predicted to have a birth 
by the end of each spell.
Abortion rates within each spell are based on the predicted number of abortions each
woman would have if she went through the entire spell.
To generate conditions such as sex composition and fecundity for second and higher spells 
each woman is randomly assigned outcomes based on the estimation results and
her characteristics as follows.
For each quarter, her probabilities of having a boy, a girl, or no birth is calculated.
She is assigned one of these outcomes based on a random draw from a uniform distribution.
The first quarter where she has a draw that results in a birth she leaves the spell
and the quarter reached is used to calculate her starting age for the subsequent spell.
If she has no draw that result in a birth within a spell her childbearing stops and
she does not enter the next spell.
Results are averaged over 1,000 repetitions.
\item[a] For individual spells the column shows the percent of women who started the
spell and ended up with a birth by the end of the spell period covered.
For the ``Overall'' row, the number is the expected number of children born per woman over 
the first four spells, conditional on having at least one birth.
\item[b] For individual spells the column shows the number of abortions per 100
births predicted to take place by the end of the spell.
For the ``Overall'' row, the number is the total number of abortion per 100 births 
predicted to take place over all four spells.
\item[c]  For individual spells the column shows the number of abortions 
predicted to take place by the end of the spell per 100 women who started the spell .
For the ``Overall'' row, the number is the total number of abortion 
predicted to take place over all four spells per 100 women who had at least one child.
\item[d] Covers first through fourth spells. 
\end{tablenotes}
\end{threeparttable}
\end{scriptsize}
\end{center}
\end{table}

The results show a substantial reduction in fertility over time.
% The percentage women with a birth by the end of each spell, conditional to making
% it to the beginning of the spell, has decreased more than
% 10 percentage points between the two periods.%
% \footnote{
% The exception is rural women's second spell, where the reductions between periods
% were 4.9 and 4.5 percentage points.
% }
Predicted completed fertility went from 2.7 for urban women married in  
the 1972--1984 period to just 2 for urban women married in the 1995--2006 period.
Hence, urban fertility is now below replacement, especially considering that
these predictions are conditional on having at least one child.
Rural women showed a similar decline in completed fertility, going from 3.2 to 2.6.
Corresponding to the decline in overall fertility, there were also substantial
reductions in the individual parity progression rates.
For example, urban women with 3 children went from having a 57\% chance of having a 
4th child for the 1972--1984 cohort to a 33\% chance for the 1995--2006 cohort---and 
they had a substantially lower probability of even making it to 3 children
in the first place.

@@

% \begin{sidewaystable}
\begin{table}[htp]
\begin{center}
\begin{scriptsize}
\begin{threeparttable}
% \caption{Descriptive Statistics by Education Level and Beginning of Spell for First and Second Spell}
\caption{Descriptive Statistics by Education Level and Beginning of Spell}
\label{tab:des_stat1}
\begin{tabular} {@{} c l D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} D{.}{.}{1.3} @{}} \toprule
                    &                     & \multicolumn{3}{c}{No Education}                                & \multicolumn{3}{c}{1--7 Years of Education}                      & \multicolumn{3}{c}{8+ Years of Education} \\ \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
                    &                     & \mco{1972--1984}     & \mco{1985--1994}     & \mco{1995--2006}     & \mco{1972--1984}     & \mco{1985--1994}     & \mco{1995--2006}     & \mco{1972--1984}     & \mco{1985--1994}     & \mco{1995--2006}     \\ % \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
\midrule
\multirow{16}{*}{\rotatebox{90}{First Spell}}                   
                    &Boy born            &       0.467         &       0.438         &       0.363         &       0.476         &       0.447         &       0.368         &       0.483         &       0.453         &       0.377         \\
                    &                    &     (0.499)         &     (0.496)         &     (0.481)         &     (0.499)         &     (0.497)         &     (0.482)         &     (0.500)         &     (0.498)         &     (0.485)         \\
                    &Girl born           &       0.427         &       0.410         &       0.352         &       0.439         &       0.429         &       0.356         &       0.454         &       0.420         &       0.353         \\
                    &                    &     (0.495)         &     (0.492)         &     (0.478)         &     (0.496)         &     (0.495)         &     (0.479)         &     (0.498)         &     (0.494)         &     (0.478)         \\
                    &Censored            &       0.106         &       0.152         &       0.284         &       0.085         &       0.124         &       0.276         &       0.063         &       0.127         &       0.270         \\
                    &                    &     (0.308)         &     (0.359)         &     (0.451)         &     (0.279)         &     (0.330)         &     (0.447)         &     (0.243)         &     (0.333)         &     (0.444)         \\
                    &Urban               &       0.181         &       0.176         &       0.200         &       0.366         &       0.334         &       0.322         &       0.690         &       0.604         &       0.561         \\
                    &                    &     (0.385)         &     (0.381)         &     (0.400)         &     (0.482)         &     (0.472)         &     (0.467)         &     (0.463)         &     (0.489)         &     (0.496)         \\
                    &Age                 &      15.854         &      16.301         &      16.871         &      16.983         &      17.470         &      17.844         &      19.516         &      20.036         &      20.766         \\
                    &                    &     (2.482)         &     (2.643)         &     (2.858)         &     (2.733)         &     (3.041)         &     (3.221)         &     (3.331)         &     (3.614)         &     (3.916)         \\
                    &Owns land           &       0.589         &       0.578         &       0.540         &       0.500         &       0.497         &       0.482         &       0.321         &       0.387         &       0.414         \\
                    &                    &     (0.492)         &     (0.494)         &     (0.498)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.467)         &     (0.487)         &     (0.493)         \\
                    &Sched.\ caste/tribe &       0.357         &       0.395         &       0.441         &       0.170         &       0.227         &       0.318         &       0.071         &       0.119         &       0.175         \\
                    &                    &     (0.479)         &     (0.489)         &     (0.497)         &     (0.376)         &     (0.419)         &     (0.466)         &     (0.257)         &     (0.324)         &     (0.380)         \\
                    &Number of quarters  & \mco{  330,315         } & \mco{  257,757         } & \mco{   75,841         } & \mco{  108,463         } & \mco{  108,863         } & \mco{   54,704         } & \mco{  102,238         } & \mco{  154,582         } & \mco{  116,272         } \\
                    &Number of women     & \mco{   30,004         } & \mco{   27,669         } & \mco{    9,684         } & \mco{   11,162         } & \mco{   13,104         } & \mco{    7,510         } & \mco{   11,271         } & \mco{   19,072         } & \mco{   16,620         } \\
                    \addlinespace
% \cmidrule(lr){1-2} \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
\midrule
\multirow{22}{*}{\rotatebox{90}{Second Spell}}
                    & Boy born            &       0.490         &       0.444         &       0.359         &       0.477         &       0.437         &       0.338         &       0.457         &       0.399         &       0.285         \\
                    &                     &     (0.500)         &     (0.497)         &     (0.480)         &     (0.500)         &     (0.496)         &     (0.473)         &     (0.498)         &     (0.490)         &     (0.451)         \\
                    & Girl born           &       0.451         &       0.413         &       0.327         &       0.462         &       0.413         &       0.319         &       0.426         &       0.351         &       0.239         \\
                    &                     &     (0.498)         &     (0.492)         &     (0.469)         &     (0.499)         &     (0.492)         &     (0.466)         &     (0.495)         &     (0.477)         &     (0.427)         \\
                    & Censored            &       0.059         &       0.143         &       0.314         &       0.061         &       0.150         &       0.344         &       0.117         &       0.251         &       0.476         \\
                    &                     &     (0.236)         &     (0.350)         &     (0.464)         &     (0.239)         &     (0.357)         &     (0.475)         &     (0.321)         &     (0.433)         &     (0.499)         \\
                    & One boy             &       0.527         &       0.515         &       0.508         &       0.519         &       0.513         &       0.505         &       0.513         &       0.520         &       0.515         \\
                    &                     &     (0.499)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         \\
                    & One girl            &       0.473         &       0.485         &       0.492         &       0.481         &       0.487         &       0.495         &       0.487         &       0.480         &       0.485         \\
                    &                     &     (0.499)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         &     (0.500)         \\
                    & Urban               &       0.181         &       0.180         &       0.212         &       0.370         &       0.348         &       0.345         &       0.697         &       0.628         &       0.578         \\
                    &                     &     (0.385)         &     (0.384)         &     (0.409)         &     (0.483)         &     (0.476)         &     (0.475)         &     (0.459)         &     (0.483)         &     (0.494)         \\
                    & Age                 &      17.774         &      18.357         &      18.870         &      18.620         &      19.200         &      19.581         &      20.988         &      21.589         &      22.252         \\
                    &                     &     (2.743)         &     (3.054)         &     (3.284)         &     (2.864)         &     (3.206)         &     (3.373)         &     (3.371)         &     (3.621)         &     (3.901)         \\
                    & First spell length  &      25.630         &      29.145         &      29.133         &      21.924         &      24.513         &      24.787         &      20.608         &      22.154         &      22.217         \\
                    &                     &    (20.345)         &    (25.044)         &    (27.216)         &    (18.142)         &    (22.027)         &    (23.204)         &    (15.371)         &    (18.651)         &    (19.176)         \\
                    & Owns land           &       0.590         &       0.572         &       0.528         &       0.497         &       0.492         &       0.468         &       0.313         &       0.369         &       0.402         \\
                    &                     &     (0.492)         &     (0.495)         &     (0.499)         &     (0.500)         &     (0.500)         &     (0.499)         &     (0.464)         &     (0.483)         &     (0.490)         \\
                    & Sched.\ caste/tribe &       0.353         &       0.390         &       0.430         &       0.164         &       0.218         &       0.304         &       0.068         &       0.110         &       0.169         \\
                    &                     &     (0.478)         &     (0.488)         &     (0.495)         &     (0.371)         &     (0.413)         &     (0.460)         &     (0.251)         &     (0.313)         &     (0.374)         \\
                    & Number of quarters  & \mco{  176,227         } & \mco{  231,546         } & \mco{   68,455         } & \mco{   67,432         } & \mco{  103,656         } & \mco{   49,236         } & \mco{   77,416         } & \mco{  164,752         } & \mco{  119,864         } \\
                    & Number of women     & \mco{   21,171         } & \mco{   28,251         } & \mco{    9,428         } & \mco{    8,209         } & \mco{   12,449         } & \mco{    6,542         } & \mco{    8,204         } & \mco{   16,601         } & \mco{   13,610         } \\
\bottomrule
\end{tabular}
\begin{tablenotes} \tiny
% \item \hspace*{-7pt} \textbf{Note.}
\item \hspace*{-0.7em} \textbf{Note.}
Means without parentheses and standard deviation in parentheses.
Interactions between variables, baseline hazard dummies and squares not shown.
Quarters refer to number of 3 month periods observed.
\end{tablenotes}

>>>>

The males-to-females ratio at birth increased dramatically over the last three 
decades in India as access to prenatal sex determination expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth \citep{Yi1993,park95}.
}
Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask about use of 
sex selection show signs of serious under-reporting \citep{goodkind96}.

With no direct information, researchers have relied on a simple method for establishing 
factors that affect sex selection: use the sex of children born as the 
dependent variable, and estimate the effects of variables on the probability of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

@@

Recent research suggests that son preference in India, when measured as ideally having 
more boys than girls, is decreasing over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
In NFHS, and Demographic and Health Surveys in general, son preference is measured as part 
of the general fertility questions.
The first question is how many children a woman would have if she could choose exactly how 
many.
This if followed by the sex preference question:
``How many of these children would you like to be boys, how many would you like to be girls 
and for how many would the sex not matter?''
}

@@

Lowering the cost of prenatal sex determination generally leads to 
a smaller percentage of parents who change their use of sex selection.
But, the total number of spells with a change may still
go up because lower cost of prenatal sex determination also
increases the use of sex selection.

@@

% What are the model highlights 
There are three main conclusions from the model.
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

@@

As the discussion above shows there are three main
implications of the availability of sex-selective abortions
combined with son preference:
a higher probability that next child is a son, 
longer spacing to next birth with use of sex selection, 
and that the use of prenatal sex determination may change between 
one observed birth and the next.
All three must be incorporated into the estimation strategy
to provide unbiased and precise estimates of the use of sex-selective abortions.%

To understand son preference when sex selection is available,
we need an empirical model that can jointly estimate the sex of
children born and the spacing between birth.

@@

[This need a much better explanation; the survival curve captures the spacing
between births and how long it takes; may need to combine with and without
sons to show differences in spacing over time; maybe it should also have
predicted average/median spacing with some confidence intervals to test]

@@

The data come from the three rounds of the National Family Health Survey 
(NFHS-1, NFHS-2 and NFHS-3),
collected in 1992--93, 1998--99, and 2005--2006.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has a small number of observation 
collected in 2000.
}
The International Institute for Population Sciences in Mumbai collected the data,
which have nationwide coverage.
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households,
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households
and NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households.

@@

By dividing spells into these periods I can examine how the use of sex selection 
has changed across the three different regimes.
Note that the periods are based on the spells' beginning year, and some spells 
will therefore cover two periods.
A couple may, for example, be married in 1984, but not have their first child until 1986.
That couple's first spell will be in the 1972--1984 period, even though most of the 
spell actually falls in the 1985--1994 period.
Some children born from spells that began in the 1972--1984 period may therefore have been
conceive when prenatal sex determination techniques were available, which could result
in evidence of sex-selective abortions even for this period.
Similarly, a spell that began in the 1985--1994 period may have been partly or mostly
under the PDNT act.
The overall effect is to bias downwards any differences between the periods.

@@

@@

The extent to which people use sex selection for different sex compositions
of previous children is of interest for two reasons.
First and foremost, it can help us understand what form son preference takes in India.
This is important in itself, but is especially important as we try to predict 
what will happen to sex selection use in the future.
[need an explanation of this based on what I have in the introduction]
Second,
if families use sex selection to reach two or more sons, progression rates for families 
with one son may overestimate the decline in fertility.

@@

% Tables of predictions 
\input{../tables/predictedFertility.tex}

The results show a substantial reduction in fertility over time.
Predicted completed fertility went from 2.7 for urban women married in  
the 1972--1984 period to just 2 for urban women married in the 1995--2006 period.
Hence, urban fertility is now below replacement, especially considering that
these predictions are conditional on having at least one child.
Rural women showed a similar decline in completed fertility, going from 3.2 to 2.6.
Corresponding to the decline in overall fertility, there were also substantial
reductions in the individual parity progression rates.
For example, urban women with 3 children went from having a 57\% chance of having a 
4th child for the 1972--1984 cohort to a 33\% chance for the 1995--2006 cohort---and 
they had a substantially lower probability of even making it to 3 children
in the first place.

@@

% Descriptive statistics tables
\input{../tables/desStat.tex}
